### 2011-10-10 papers/smectics/smectics.tex
Model part and 3D-result part reworked
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1457 40876441-3496-0410-a613-a2fca35b2ae2

\twocolumn[
  \begin{@twocolumnfalse}
\noindent\LARGE{\textbf{Nonlinear Rheology of Layered Smectic Liquid Crystals}}
\vspace{0.6cm}

\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{b}$}, K. Stratford\textit{$^{c}$} and M.E. Cates \textit{$^{b}$}}
}\vspace{0.5cm}
%Please note that \ast indicates the corresponding author(s) but no footnote text is required.

@@

\noindent \normalsize{In this work we present large scale computer simulations of the bulk rheology of a smectic liquid crystal at intermediate to large values of the shear flow, threfore in the nonlinear regime. In two dimensions we find that, according to the thermodynamic parameters, shear may either align the system and stabilise an almost regular lamellar phase, or induce the nucleation and proliferation of defects in the smectic layers. In this last case, defects may be progressively annealed by slowly decreasing the applied shear rate, so that there are marked memory effects and the rheology is history-dependent. Finally, selected simulations in three dimensions show shear-induced ordering rather than defect nucleation, thus pointing out to a dramatic effect of dimensionality on the physics of the system.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

\footnotetext{\textit{$^{a}$~University College of London, London, UK}}
\footnotetext{\textit{$^{b}$~SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
\footnotetext{\textit{$^{c}$~EPCC, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}

@@

\section{Introduction}
Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurth etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid. Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibers constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character. Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in biosensing.

@@

The rheological properties of a non-Newtonian fluid are often linked to the microstrucure of the material. For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate. Another intriguing example of strongly non-Newtonian rheology is region I flow in polymeric nematic liquid crystals. While our understanding of the physics of region I flow is far from complete, it is thought that a defect network may form and lead to much enhanced viscosity, with marked shear thinning, $\eta \sim \dot{\gamma}^{\alpha}$, and an exponent $\alpha$ typically between 0 and 1. Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics. Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via preshearing, or not. In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample. An intriguing flow mode which is possible in layered system is permeation, through which the individual molecules in the fluid displace keeping the defect pattern intact in steady state. This leads to increased dissipation and hence to a larger viscosity. A well-known example of permeation flow is the one which occurs e.g. in cholesterics which are sheared or pushed along the direction of the cholesteric helix. However, one would expect permeation flows to be possible also when a non-translationally invariant defect pattern is present in the system, so that it may occur to some extent also in quenched smectics for instance.
Lamellar rheology is also of interest for a number of additional intriguing phenomenological observations beyond their nonlinear shear-stress curve. For instance, in block-copolymers one may observe, under some conditions, shear-induced phase transitions from the microphase separated state to ordered phases like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}, and it would be desirable to extend our theoretical understanding of such switches. Furthermore, lyotropic layered liquid crystals have sometimes been observed to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93} -- the mechanism underlying such pattern formation has remained elusive to date.

Despite the wealth of experimentally available results, and the importance of rheology to for instance industrial applications, flow behaviour remains challenging to study theoretically. This is mainly in view of the highly non-linear equations of motion which require either drastic approximations or challenging computer simulations. At the same time, realistic modelling eventually requires the simulation of a fully 3-dimensional flow, which is particularly expensive computationally. In this work, we aim to contribute to fill in the gap towards realistic large scale computational studies of complex fluids, and our focus is the physics of smectics under shear. For another example of large scale simulations for complex fluids, see Ref.~\cite{Saksena09} where sheared gyroid phases were simulated.
We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters. In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like miscelles, surfactants, smectic liquid crystals and block-copolymer.

%Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organize in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.

@@

The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimizes the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.
\begin{eqnarray}
{\cal F}[\phi]&=&\int d^dx f(\phi({\bm x}))\nonumber\\
&=&\int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\bm \nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\label{FE-functional}
\end{eqnarray}
The parameters $b$ and $c$ are positive to ensure stability. 
For $a>0$ there is only one solution $\phi\equiv0$ which minimizes the functional and corresponds to the disordered state.
For $a<0$ a microphase separation occurs and depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) because the formation of interfaces become energetically favourable.\\
The dynamics of the binary fluid is governed by a convection-diffusion equation with a diffusive current proportional to the chemical potential $\mu$, which is the functional derivative of ${\cal F}$ with respect to $\phi$.
\begin{equation}\label{OPEOM}
\partial_t \phi +{\bm \nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}
The coefficient $M$ herein plays the role of a mobility coefficient.\\
If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys a Navier-Stokes equation
\begin{equation}\label{NSE}
\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha = \partial_\beta \sigma_{\alpha \beta}. 
\end{equation}
The stress tensor reads
\begin{equation}
\sigma_{\alpha \beta}=-P^{th}_{\alpha \beta} +\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha).
\end{equation}
It consists of a hydrodynamic part, which is proportonial to the viscosity $\eta$ of the solvent and a thermodynamical part $P^{th}$ that depends on the order parameter.
The latter can be calculated from the free energy functional:
\begin{equation}
P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + D_{\alpha \beta}(\phi).
\end{equation}  
In this expression the symmetric tensor $D_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee the mechanical equilibrium \cite{Evans79}.
The thermodynamic pressure tensor is given by
\begin{equation}
P^{th}_{\alpha \beta}=p\, \delta_{\alpha \beta} + P^{chem}_{\alpha \beta}
\end{equation}
with an isotropic pressure contribution
\begin{eqnarray}
p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\bm \nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}
and an anisotropic part
\begin{equation}
P^{chem}_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}

The Navier-Stokes Eq.\ref{NSE} was solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
The velocity field were then put into the convection-diffusion equation Eq.\ref{OPEOM} and solved at the same timestep. 
We assumed periodic boundary conditions along the flow and vorticity direction and Lees-Edwards boundary conditions (BC) \cite{Wagner02} along the direction of the velocity gradient.
The total number of Lees-Edwards planes was varied according to the size of the simulation so that the distance of two planes was always 16 lattice sites.
To impose simple shear flow, the planes in the upper half of the velocity gradient direction were moved at a constant speed in positive flow direction, whereas the planes in the lower hemisphere were moved in negative flow direction.\\
Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as
\begin{equation}
C({\mathbf k})=\phi({\mathbf k})\phi(-{\mathbf k}),
\end{equation}
with $\phi({\mathbf k})$ as normalised Fourier-transform of the order parameter.
The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
In order to analyse the relation between rheological quantities and the morphology of the system it was necessary to quantify the number of defects.
For this purpose we defined an alignment factor $X_i$, which measures the average projection of the normalised order parameter gradients onto the gradient at site $i$.
More explicitely the alignment factor $X_i$ is given as
\begin{equation}\label{alignment}
X_i=\frac{1}{N_{nn}} \sum_{j=1,N_{nn}}\left(\frac{\nabla \phi_i \cdot \nabla \phi_j}{|\nabla\phi_i||\nabla\phi_j|}\right)^2. 
\end{equation}
$N_{nn}$ is the number of nearest neighbours ($N_{nn}=8$ in 2D and $N_{nn}=26$ in 3D).
For smooth alignment and small curvature of the lamellae $X_i$ is very close to unity.
At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes up smaller positive values.
By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates from what would be recognised by the eye as an aligned configuration.
We found a threshold of $\xi=0.8$ to work quite well for our requirements.
Unfortunately this defect detecting protocol is not able to assign only a single defect site to a disturbed region in the order parameter structure as normally more than one site close to a defect would have an alignment factor below the threshold.
However, as the number of sites per defect is roughly constant throughout the sample, it is still useful for quantifying the degree of alignment and comparing different systems or flow conditions.\\
By relating the number of defects sites with $X_i$ below the threshold to the total number of sites it is possible to define a defect density:
\begin{equation}\label{defectdensity}
\rho_D=\frac{1}{N} \sum_{i=1,N} \theta(\xi-X_i).  
\end{equation}
The function $\theta$ is the Heaviside function and $N=N_x\times N_y \times N_z$ the total number of sites in the simulation.
The exact value of the defect density depends clearly on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.

@@

In the following we describe how the simulation units can be mapped onto real physical units.
It is necessary to calibrate the scales for mass density, length and time or alternatively as we do for mass density, length and energy density.\\
As usual we set the lattice spacing $l$, time step $\Delta t$ and mass density of the reference fluid $\rho_0$ all to unity in LB units (LBU).
It is straightforward to determine the unit of mass. 
A typical block copolymers or liquid crystal has a mass density the region of $10^3 {\rm kg m}^{-3}$, which is of the order ${\cal O}(1)$ in LBU.
Hence $1 {\rm kg m}^{-3}=10^{-3}$ LBU.\\
In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig12} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space.\\
The energy scale is slightly more difficult to calibrate.
We use the fact that both diblock copolymer melts and smectic liquid crystals belong to the Brazovskii class.
This has been already pointed out by Leibler \cite{Leibler80} in his seminal work on microphase separation in block copolymers.
The Landau free energy functional we are using can be transformed into a Brazovskii-type of form and coefficients can be compared \cite{Xu05, Fredrickson89}.
The energy density scale in our functional \ref{FE-functional} is set by the coefficients $-a=b$ of the quadratic and quartic term, respectively.
Referring to \cite{Fredrickson89} we can derive an expression that relates $b$ to Leibler's original parameters.
This formula reads 
\begin{equation}
b=\frac{u_0 k_B T}{6}=\frac{c^4\lambda k_B T}{6} \rho_c,
\end{equation}
whereas $\rho_c$ is the number density of chains per unit volume,
often expressed as $\rho_c=\sqrt{N/6^3} R_g^{-3}$ with $N$ as roughly the number of Kuhn segments and $R_g$ as radius of gyration.
Typical values are $N\simeq10^4$ and $R_g\simeq 100{\rm nm}$.
With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
Hence the LB unit of energy density is about $1\e{6} \rm{Pa}$.\\
We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=1\e{-3+6+14} \rm{LBU}\simeq1\e{17} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq3\e{8}$LBU and one LB timestep $\Delta t$ corresponds to roughly 3ns.\\
Finally we want to get an estimated scale for shear rates. 
A common way to quantify them is the dimensionless Peclet number, which measures the transport of a particle as the ratio of the rate of advection to the rate of diffusion.
The Peclet number can be defined as $Pe_0=\gd\, D^2/D_0$, where $\gd$ is the rate of strain, $D$ is a characteristic length, e.g. the diameter of a particle and $D_0$ is the diffusivity.
The highest shear rate we apply is $2\e{-4}$ in LBU, basically reciprocal LB time units.
With the above value for $\Delta t$ we can convert this into about $5\e{4}{\rm s}^{-1}$.
If we suppose the Stokes-Einstein diffusion constant is a good approximation for $D_0$, we get for an object of the size of $D\simeq5\e{-8}$ at room temperature in a low viscous solevent like water values of about $D_0=1\e{-11} \rm{m}^2\rm{s}^{-1}$.
This results in Peclet number of the order ${\cal O}(10)$ and means shear rates $\gd={\cal O}(10^{-4})$ in LBU correspond to a Peclet numbers $Pe_0={\cal O}(10)$.

@@

Our discussion will be in two parts, beginning with simulations in two dimensions for a system A and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
The stiffness of the system was controlled by modifying the Landau parameters $a$ and $b$ of the bulk terms in the free energy functional \ref{FE-functional} with the parameters $\kappa$ and $c$ multiplying the gradient terms held constant.
By this we obtained slightly different segregation levels in both systems A and B, whereas the lamellar distance was virtually the same.\\
The second part of our discussion consits of simulations in three dimensions using identical parameters to system A and a box size $N_x\times N_y\times N_z=128\times256\times128$.
We will refer to this system as system C.
Table \ref{tab1} at the end of this paragraph provides an overview of all simulation parameters.

\subsection*{Results in Two Dimensions}
There is some freedom regarding the parametrizaton of binary fluids using the Landau-Brazovskii free energy approach \ref{FE-functional}.
Hence we stuck in the present work to parameters, which have been found to cover the relevant morphology of lamellar systems and produce stable outcomes in previous studies \cite{Kendon01,Xu03, Xu06b}. 
We performed simulations for a variety of parameter sets with Landau parameters $1\cdot10^{-3}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
Smaller Landau parameters slowed down the ripening speed of the lamellar structure, which is consistent with the fact that the system is closer to the critical point where the phase microphase separation occurs.
Higher viscosities led to shorter interfaces, whereas values around the lower threshold invoked longer interfaces.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_run703_320.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_500.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
\caption{Order parameter $\phi$ of system A: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom). Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
\label{fig1}
\end{figure}

>>>>

\twocolumn[
  \begin{@twocolumnfalse}
\noindent\LARGE{\textbf{Large-scale Simulation of Layered Binary Fluids and Smectic Liquid Crystals in Shear Flow}}
\vspace{0.6cm}

\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.V. Coveney{$^{a}$} and M.E. Cates \textit{$^{c}$}}
}\vspace{0.5cm}
%Please note that \ast indicates the corresponding author(s) but no footnote text is required.

@@

\noindent \normalsize{In this work we present large scale computer simulations of the bulk rheology of a smectic liquid crystal at intermediate to large values of the shear flow, therefore in the nonlinear regime. In two dimensions we find that, according to the thermodynamic parameters, shear may either align the system and stabilise an almost regular lamellar phase, or induce the nucleation and proliferation of defects in the smectic layers. In this last case, defects may be progressively annealed by slowly decreasing the applied shear rate, so that there are marked memory effects and the rheology is history-dependent. Finally, selected simulations in three dimensions show shear-induced ordering rather than defect nucleation, thus pointing out to a dramatic effect of dimensionality on the physics of the system.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, UK}}
\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK}}
\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, UK}}

@@

\section{Introduction}
Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurt etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid. Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibres constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character. Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in bio-sensing.

@@

The rheological properties of a non-Newtonian fluid are often linked to the microstructure of the material. For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate. Another intriguing example of strongly non-Newtonian rheology is region I flow in polymeric nematic liquid crystals. While our understanding of the physics of region I flow is far from complete, it is thought that a defect network may form and lead to much enhanced viscosity, with marked shear thinning, $\eta \sim \dot{\gamma}^{\alpha}$, and an exponent $\alpha$ typically between 0 and 1. Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics. Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via pre-shearing, or not. In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample. An intriguing flow mode which is possible in layered system is permeation, through which the individual molecules in the fluid displace keeping the defect pattern intact in steady state. This leads to increased dissipation and hence to a larger viscosity. A well-known example of permeation flow is the one which occurs e.g. in cholesterics which are sheared or pushed along the direction of the cholesteric helix. However, one would expect permeation flows to be possible also when a non-translationally invariant defect pattern is present in the system, so that it may occur to some extent also in quenched smectics for instance.
Lamellar rheology is also of interest for a number of additional intriguing phenomenological observations beyond their nonlinear shear-stress curve. For instance, in block-copolymers one may observe, under some conditions, shear-induced phase transitions from the microphase separated state to ordered phases like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}, and it would be desirable to extend our theoretical understanding of such switches. Furthermore, lyotropic layered liquid crystals have sometimes been observed to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93} -- the mechanism underlying such pattern formation has remained elusive to date.

Despite the wealth of experimentally available results, and the importance of rheology to for instance industrial applications, flow behaviour remains challenging to study theoretically. This is mainly in view of the highly non-linear equations of motion which require either drastic approximations or challenging computer simulations. At the same time, realistic modelling eventually requires the simulation of a fully 3-dimensional flow, which is particularly expensive computationally. In this work, we aim to contribute to fill in the gap towards realistic large scale computational studies of complex fluids, and our focus is the physics of smectics under shear. For another example of large scale simulations for complex fluids, see Ref.~\cite{Saksena09} where sheared gyroid phases were simulated.
We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters. In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like micelles, surfactants, smectic liquid crystals and block-copolymer.

%Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organise in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.

@@

The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.

\begin{eqnarray}
{\cal F}[\phi]&=&\int d^dx f(\phi({\bm x}))\nonumber\\
&=&\int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\bm \nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\label{FE-functional}
\end{eqnarray}

In order to ensure thermodynamic stability the parameters $b$ and $c$ have to be positive.
For $a>0$ there is only one unique solution $\phi\equiv0$ which minimises the functional Eq. \ref{FE-functional} and wich reprsents the the disordered state.
For $a<0$ a microphase separation occurs.
Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$, whereas the interfacial width is roughly $\sqrt{2|\kappa|/|a|}$. 
This entails that by varying $a$ (and $b\sim |a|$) by about on decade, the segregation of the two phases varies by about a factor $3$.\\
The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.

\begin{equation}\label{OPEOM}
\partial_t \phi +{\bm \nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}

The coefficient $M$ herein plays the role of a mobility coefficient and the diffusivity of the system can be approximated by $D_0\simeq M\;b$.\\
If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation

\begin{equation}\label{NSE}
\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha = \partial_\beta \sigma_{\alpha \beta}. 
\end{equation}

The stress tensor in this equation reads

\begin{equation}
\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-P^{th}_{\alpha \beta}.
\end{equation}

It consists of a hydrodynamic part, which is proportional to the background viscosity $\eta$ and velocity gradients and a thermodynamical part $P^{th}$ that depends on the order parameter and gradients thereof:

\begin{equation}
P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + {\cal D}_{\alpha \beta}(\phi).
\end{equation}

In this last expression the symmetric tensor ${\cal D}_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee mechanical equilibrium \cite{Evans79}.
The thermodynamic pressure tensor is given by

\begin{equation}
P^{th}_{\alpha \beta}=p\, \delta_{\alpha \beta} + P^{chem}_{\alpha \beta}
\end{equation}

with an isotropic pressure contribution

\begin{eqnarray}
p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\bm \nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}

and an anisotropic part

\begin{equation}
P^{chem}_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}

The Navier-Stokes Eq.\ref{NSE} was solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
The velocity field was then used in the advection-diffusion equation Eq.\ref{OPEOM}, which was solved by a finite-difference method at the same timestep. 
We assumed periodic boundary conditions along the flow and vorticity direction and Lees-Edwards boundary conditions (BC) \cite{Wagner02} along the direction of the velocity gradient.
The total number of Lees-Edwards planes was varied according to the size of the simulation so that the distance of two planes was always 16 lattice sites.
To impose simple shear flow, the planes in the upper half of the velocity gradient direction were moved at a constant speed in positive flow direction, whereas the planes in the lower hemisphere were moved in negative flow direction.\\
Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as

\begin{equation}
C({\mathbf k})=\phi({\mathbf k})\phi(-{\mathbf k}),
\end{equation}

with $\phi({\mathbf k})$ being the normalised Fourier-transform of the order parameter.
The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
For the analysis of the relation between rheological quantities and the morphology of the system it was necessary to find a means to quantify the number of defects.
For this purpose we defined an alignment factor $X_i$, which measures the average projection of the normalised order parameter gradients onto the gradient at site $i$.
More explicitly the alignment factor $X_i$ is given by

\begin{equation}\label{alignment}
X_i=\frac{1}{N_{nn}} \sum_{j=1,N_{nn}}\left(\frac{\nabla \phi_i \cdot \nabla \phi_j}{|\nabla\phi_i||\nabla\phi_j|}\right)^2. 
\end{equation}

$N_{nn}$ is the number of nearest neighbours ($N_{nn}=8$ in 2D and $N_{nn}=26$ in 3D).
For smooth alignment and small curvature of the lamellae $X_i$ is very close to unity.
At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes smaller positive values.
By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates preceivably from what would be recognised as an aligned configuration exhibiting lamellar order.
We found a threshold value of $\xi=0.8$ to be well suitable for our requirements.
Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment factor below the threshold.
However, as the number of sites per defect is on average more or less constant throughout the sample, this protocol provides still a useful way to quantify the degree of order and alignment in various flow conditions.\\
By normalising the number of defects sites with $X_i$ below the threshold a defect density can be defined:

\begin{equation}\label{defect density}
\rho_D=\frac{1}{N} \sum_{i=1,N} \theta(\xi-X_i).  
\end{equation}

The function $\theta$ is the Heaviside function and $N=N_x\times N_y \times N_z$ the total number of sites in the simulation.
The exact value of the defect density clearly depends on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.

@@

In the following we describe how the simulation units can be mapped onto real physical units.
It is necessary to calibrate the scales for mass density, length and time or alternatively as we do for mass density, length and energy density.\\
As usual we set the lattice spacing $l$, time step $\Delta t$ and mass density of the reference fluid $\rho_0$ all to unity in LB units (LBU).
It is straightforward to determine the unit of mass. 
A typical block copolymers or liquid crystal has a mass density the region of $10^3 {\rm kg m}^{-3}$, which is of the order ${\cal O}(1)$ in LBU.
Hence $1 {\rm kg m}^{-3}=10^{-3}$ LBU.\\
In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig12} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space or $1 {\rm m}=2\cdot10^8$ LBU.\\
The energy scale is slightly more difficult to calibrate.
We use the fact that both diblock copolymer melts and smectic liquid crystals belong to the Brazovskii class.
This has been already pointed out by Leibler \cite{Leibler80} in his seminal work on microphase separation in block copolymers.
The Landau free energy functional we are using can be transformed into a Brazovskii-type of form and coefficients can be compared \cite{Xu05, Fredrickson89}.
The energy density scale in our functional \ref{FE-functional} is set by the coefficients $-a=b$ of the quadratic and quartic term, respectively.
Referring to \cite{Fredrickson89} we can derive an expression that relates $b$ to Leibler's original parameters.
This formula reads 
\begin{equation}
b=\frac{u_0 k_B T}{6}=\frac{c^4\lambda k_B T}{6} \rho_c,
\end{equation}
whereas $\rho_c$ is the number density of chains per unit volume,
often expressed as $\rho_c=\sqrt{N/6^3} R_g^{-3}$ with $N$ as roughly the number of Kuhn segments and $R_g$ as radius of gyration.
Typical values are $N\simeq10^4$ and $R_g\simeq 100{\rm nm}$.
With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
Hence the LB unit of energy density is about $10^{6} \rm{Pa}$ or $1\rm{Pa}=10^{-6}$ LBU.\\
We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=(1\e{-3}) \cdot (1\e{6}) \cdot (4\e{16})\rm{LBU}\simeq4\e{19} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq6.3\e{9}$ LBU and one LB timestep $\Delta t$ corresponds to roughly $0.15 \rm{ns}$.\\
Finally we want to get an estimated scale for shear rates. 
A common way is to express the ratio of transport via advection to that via diffusion by means of the Peclet number.
The Peclet number can be defined as $Pe=\gd\, L^2/D_0$, whereas $\gd$ is the strain rate, $L$ is a characteristic length scale, e.g. the lamellar width and $D_0$ is the diffusivity.
Because the Peclet number is a dimensionless quantity, we can perform this estimation conveniently in lattice units.
The highest shear rate we apply are of the order $1\e{-4}$ in reciprocal LB time units.
With the above given values for the lamellar width of $L\simeq10$ and diffusivity of system A $D_0=M\;b\simeq 0.25 \cdot 5\e{-4}=1.25\e{-4}$ this results in Peclet numbers $Pe=(1\e{-4})\cdot(1\e{2})\cdot(1.25\e{-4})^{-1}=80={\cal O}(10^2)$.
Hence, the conversion between shear rates and Peclet numbers is roughly $\gd=10^{-4}\cong Pe=10^2$.

@@

Our discussion will be in two parts, beginning with simulations in two dimensions for a system A-2D and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
The stiffness of the system was controlled by modifying the Landau parameters $a$ and $b$ of the bulk terms in the free energy functional \ref{FE-functional} with the parameters $\kappa$ and $c$ multiplying the gradient terms held constant.
By this we obtained slightly different segregation levels in both systems A and B, whereas the lamellar distance was virtually the same.\\
The second part of our discussion consists of simulations in three dimensions using identical parameters to system A-2D and a box size $N_x\times N_y\times N_z=128\times256\times128$.
We will refer to this system as system A-3D.
Table \ref{tab1} provides an overview of all simulation parameters.
